Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

### 2020-01-23 manuscript/main.Rmd
revision

abstract: |
   Land Use and Land Cover change (LULCC) of African rainforest contribute to global carbon emissions. Yet, most historical estimates of LULCC and their carbon emissions rely on non-spatially explicity data in the pre-satellite era (< 1972). Past studies have expanded this pre-satellite time frame with declassified satellite surveillance data or aerial surveys, but none cover the Congo Basin. Here, we use historical aerial survey photos to map the extent and structure of LULCC around Yangambi, central Congo Basin, in 1958. Our study leveraged Structure-from-Motion to generate a large orthomosaic covering 828 km^2^, counting 733 million pixels, at a sub meter resolution and geo-referenced to ~4.7 $\pm$ 4.3 m. Primary forest in the orthomosaic was classified with a Convolutional Neural Network Deep Learning approach. We used these data to quantify LULCC, landscape and canopy complexity and scale above ground biomass between historical and contemporary condition. We show a shift from historical highly structured industrial deforestation in 1958 (162 km^2^), to contemporary smallholder farming clearing (88 km^2^), increasing landscape fragmentation and forest edges. Efforts to quantify canopy texture and their link to carbon storage had limited to no success. Our analysis provides an insight in key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale and provide a historical context to land-cover and land-use change spatial patterns for past and ongoing field research in the area.
keywords: |
  Aerial survey, data recovery, CNN, Deep Learning, SfM, Congo Basin
acknowledgement: |
  This research was supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE). KH acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668). I'm grateful for the support of Thales D'Haulleville volunteering his time in scanning the images.
authorcontributions: |
  K.H. conceived and designed the study, analyzed the data, prepared figures and tables, authored, reviewed and approved the final draft of the manuscript. T.d.H. scanned all image data. E.K. and T.d.H. provided plot based AGC estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A., J.V.D.B., H.V and L.W. reviewed the final manuscript.
conflictsofinterest: |
  The authors declare no conflict of interest. The founding sponsors had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, an in the decision to publish the results.
output: rticles::mdpi_article
---

```{r setup, include=FALSE, message=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  library(orthodrc)
  library(raster)
  library(ggplot2)
  library(ggthemes)
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(landscapemetrics)

  # set kable NA values to blank
  options(knitr.kable.NA = '-')

@@

  row.names(change_stats) <- c(
   "Forest",
   "  -  of which edges",
   "Regrowth > 1958 loss",
   "Loss > 2000",
   "Loss > 1958 (permanent)")
```

@@

The functioning of a tropical forest, and its capacity to sequester carbon, is heavily influenced by anthropogenic disturbances such as selective logging, clear felling, or slash and burn practices, all increasing forest fragmentation. This increased pace in the loss of rainforest cover [@butsic2015] is predicted to negatively impact the function of these tropical forests, and their future capacity to sequester carbon, as anthropogenic disturbances such as selective logging, clear felling, or slash and burn practices collectively contribute to increasing forest fragmentation.

These forest fragments and their edges contain fewer large trees [@laurance2000], in part due to their hotter and brighter microclimates [@magnago2015]. Similarly, species composition and biodiversity [@poorter2006] are negatively affected. Spatial disturbances also have an explicit temporal component, where the time since disturbance affects the strength of the effects [@didham1999]. Deforestation and gap formation also influence the overall canopy structure, where pioneer species grow taller and with narrower crowns [@poorter2006; @peres2006] affecting the forest light regime and carbon dynamics [@stark2012]. As such, anthropogenic factors through land-use and land cover change (LULCC) heavily affect forest structure and functioning [@fauset2017; @brinck2017; @didham1999]. In the tropics the majority of emissions originate from deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015] contributing to 10-15% of the total global emissions [@vanderwerf2009] and concomitant losses in biodiversity losses [@barlow2016; @vandeperre2018] and how LULCC is evolving over time. Although the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015]. Yet, the tropics is now becoming a large source of CO_2 to the atmosphere as deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015] contribute up to 10-15% of the total global CO2 emissions [@vanderwerf2009]. Furthermore, recent estimates show that 31% of recent carbon emissions are caused by edge effects alone [@brinck2017]. In this context, there are very few long term forest inventory plots that presently exist in the central Congo Basin [@Verbeeck2011a; @lewis2013]. This further limits the accuracy of current biomass estimates and challenges our understanding of how forest structure and function changes in the short and long term in these ecosystems [@Kearsley2013; @Doetterl2015].

Historical estimates of LULCC, and their carbon emissions, exist [@houghton2012; @ramankutty1999; @defries2002; @sader1988], however they generally rely on non-spatially explicit data in the pre-satellite era (< 1972) [@houghton2010]. These analysis often use FAO statistics, historical land-use sources, and/or population and socio-economic data [@houghton2012], as limited earlier spatial data sources are available [@hansen2013; @achard2014]. For example, the @houghton2010 long-term study used FAO and national censuses alongside a bookkeeping model to estimate carbon emissions from LULCC. Other studies use remote sensing data, but are limited in time to recent decades [@defries2002; @houghton2012]. These proxy based studies have a limited view on long-term direct and indirect spatial effects of deforestation. As such, both the spatial structure and the fragmentation pattern of a forest combined with its life history has a profound effect on carbon emssisions. However, in this context remote sensing assessments provide more accurate spatial information to help estimate carbon emissions [@mitchard2018].

Past studies expanded knowledge during the pre-satellite time frame through the use of now declassified satellite surveillance data to map deforestation in the US, central Brazil, and post world war II Germany [@song2015; @nita2018]. In some cases these records could extend back in time even further using historical aerial survey images. In northern Europe especially these data are well known and used, for example to map the extent of forest cover across ~100 km^2^ regularly, and at times using automated workflows [@giordano2018]. Survey data across the African continent is less common or inaccessible or both. Some studies do exist, both @buitenwerf2012 and @okeke2006 used aerial survey images to map vegetation changes in South African savannas and the Adulam Nature Reserve, respectively. Whilst @frankl2015 used 20 aerial survey images to map the Ethiopian highlands of 1935. High resolution aerial or surveillance satellite images provide scientists tools to monitor forest extent and structure over long time frames. Moreover, canopy texture from aerial survey images has been successfully linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010]. Thus using texture based metrics obtained from (high resolution) remote sensing data opens up possibilities to link historical aerial footage to carbon density across large spatial extents [@Couteron2005].

Given the impact of LULCC on the structure and functioning of tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover in the pre-satellite era is an imperative [@willcock2016]. In this study we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi, in the central Congo Basin at the beginning of 1958 and the start of the antrophocene [@lewis2015]. Our analysis provides a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity. Our analysis further contextualises the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin.

@@

Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville). Black-and-White infrared images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between 9 - 11h local time. Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negative of 180 by 180 mm. Flights were flown at an average absolute altitude of ~5200 m above sea level, covering roughly 18 530 km^2^ at an approximate scale of 1/40 000. The use of the integrated autograph system ensured timely acquisition of pictures with a precise overlap (~1/3) between images. This large overlap between images together with flight parameters would allow post-processing, using stereographs, to create accurate topographic maps. Original data from this campaign are stored in the [Royal Museum for Central Africa](https://www.africamuseum.be/en) in Tervuren, Belgium (Figure 1).

```{r echo=FALSE, out.width='100%', fig.cap="A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river."}
knitr::include_graphics('./figures/print_examples.jpg')
```

@@

```{r echo=FALSE, out.width='100%', message=FALSE, warning=FALSE, fig.cap="Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study. The bounding box of the orthomosaic data presented in this study is shown as a rectangle (23x36 km). The outline of a recent high-resolution Geo-eye panchromatic image is shown as a dashed dark grey rectangle (10x10km). The location of various permanent sampling plots are shown as x, +, and open squares and triangles for the mixed, mono-dominant and edge plots respectively. The grey polygon delineates the current day Yangambi Man and Biosphere reserve. The inset, top right, situates the greater Yangabmi region with the DR Congo. The full flight plan and details are shown in Appendix Figure 1 and 2"}
knitr::include_graphics('./figures/flight_paths.png')
```

@@

All selected images, covering the Yangambi area, were contact prints as original negatives of the prints were not available. Images were scanned at a resolution exceeding their original resolution (or grain) at the maximal physical resolution of an Epson A3 flatbed scanner (i.e. 2400 dpi or 160MP per image) and saved as lossless tiff images. Data were normalized using contrast limited histogram equalization [@zuiderveld1994] with a window size of 32 and a clip limit of 1.5. Fiduciary marks were used to rectify and downsample the images into square 7700x7700 pixel images (~1200 dpi, 81MP). This resulted in a dataset with digital images at a resolution that remained above the visible grain of the photographs, whilst the reduced image size facilitated easier file handling and processing speed.

@@

We calculated the orthomosaic using a low resolution point cloud and digital elevation map (DEM). Additional ground control points were provided to assist in the referencing of image and constrain the optimization routine used in the SfM algorithm. Ground control points consisted of rooftop edges of permanent structures which could be verified in both old and new aerial imagery (i.e. ESRI World Imagery). Although clouds were removed during the SfM routine we did not mask all clouds in the final orthomosaic to maximize forest coverage. The final scene was cropped, to provide consistent wall to wall coverage of the reconstructed scene. The orthomosaic was exported as a geotiff for further georeferencing in QGIS [@qgisdevelopmentteam2018] using the georeferencer plugin (version 3.1.9) and additional ESRI World Imagery high resolution reference data. We used 3^th^ degree polynomial and 16 ground control points to correct the final image. Ground control points, raw image data and final processed image are provided in addition to measures of uncertainty such as mean root mean squared (RMSE), mean and median error across all ground control points. All subsequent analysis are executed on the final geo-referenced orthomosaic or subsets of it.

@@

### Classifying forest cover

We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands, fields and buildings. We used the Unet Convolutional Neural Net (CNN, @ronneberger2015) architecture implemented in Keras [@chollet2015] with an efficientnetb3 pre-processing backbone [@yakubovskiy2019] running on TensorFlow [@martinabadi2015] to train a binary classifier (i.e. forest or non-forested). This methodology is increasingly being used to automate pixel-level classification in (color) digital photography data [@buscombe2018]. Training data were collected from the orthomosaic by randomly selecting 513 pixel square tiles from homogeneous forested or non-forested areas within the historical orthomosaic. Homogeneous tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3). We generated 5000 synthetic landscapes for training, while 500 landscapes were generated for both the validation and the testing dataset. Source tiles did not repeat across datasets to limit overfitting. In order to limit stitch line misclassifications, along the seams of mosaicked images, we created synthetic landscapes with different forest tiles to mimick forest texture transitions. We applied this technique to 10% of the generated synthetic landscapes. The CNN model was trained for 100 epochs with a batch size of 30 using Adam optimization [@kingma2017], maximizing the Intersect-over-Union (IoU) using Sørensen–Dice and categorical cross-entropy loss functions. Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring. During final model evaluation we report the IoU of our out-of-sample test datasets. The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote across overlapping areas to limit segmentation edge effects. We refer to Figure 4 for a synoptic overview of the full deep learning learning workflow. In addition, we used the first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time) to classify and visually assess the robustness of the algorithm on contemporary remote sensing imagery with the same moving window approach.

```{r echo=FALSE, out.width='100%', fig.cap="orthomosaic", fig.cap="An example synthetic landscape, combining homogeneous forest and non-forest images into a patchy landscape using random gaussian field based masks. The left panel shows a combined synthetic landscape, while the right panel shows the corresponding forest (black) and non-forest (white) labels."}
knitr::include_graphics('./figures/synthetic_landscape.png')
```

```{r echo=FALSE, out.width='100%', fig.cap="orthomosaic", fig.cap="A diagram of the deep learning workflow followed in training a binary forest / non-forest cover convolutional neural net U-Net model to generate our forest cover map."}
knitr::include_graphics('./figures/cnn_diagram.pdf')
```

To map long term land-use and land-cover change in the Yangambi region we used the contemporary Global Forest Change version 1.6 data (GFC, tile 10N-020E) [@hansen2013]. Using this data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier. In our analysis we only included forested pixels which recorded no loss throughout the whole 2000 - 2018 period. As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30m GFC resolution. We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river. Where the datasets overlap we provide summary statistics of deforestation, reforestation and afforestation. The same methodology was used to quantify CNN model performance across the Geo-Eye 1 scene.

@@

Surveys of old plantations show a large variation in AGC, depending on age and the crop type. For example, the AGC values varied from 168.67 to 86.55 g Mg C ha^-1^, for *Hevea brasiliensis* (rubber tree) and *Elaeis guineensis* (oil palm) plots respectively (@bustillo2018, personal communications). These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see Table A1). We therefore use both the estimates of old-regrowth and mixed forest to estimate AGC for regrowth. We did not have sufficient data to account for individual changes in AGC across plantations.

@@

Hufkens et al. (2019). A curated dataset of aerial survey images over the central Congo Basin, 1958. Zenodo: doi.org/10.5281/zenodo.3547767. All data not included in the latter repository can be found bundled with the analysis code as listed below. Proprietary datasets (i.e. Geo-Eye data) are not shared, but purchase order numbers allow for acquisition of these datasets to ensure reproducibility. The @hansen2013 image data is freely available from [the GFC project website](https://earthenginepartners.appspot.com/science-2013-global-forest/download_v1.6.html).

>>>>

abstract: |
   Land Use and Land Cover change (LULCC) of African rainforest contribute to global carbon emissions. Yet, most historical estimates of LULCC and their carbon emissions rely on non-spatially explicity data in the pre-satellite era (< 1972). Past studies have expanded this pre-satellite time frame with declassified satellite surveillance data or aerial surveys, but none cover the Congo Basin. Here, we use historical aerial survey photos to map the extent and structure of LULCC around Yangambi, central Congo Basin, in 1958. Our study leveraged Structure-from-Motion to generate a large orthomosaic covering 82800 ha, counting 733 million pixels, at a sub meter resolution and geo-referenced to ~4.7 $\pm$ 4.3 m. Primary forest in the orthomosaic was classified with a Convolutional Neural Network Deep Learning approach. We used these data to quantify LULCC, landscape and canopy complexity and scale above ground biomass between historical and contemporary condition. We show a shift from historical highly structured industrial deforestation in 1958 (16200 ha), to contemporary smallholder farming clearing (8800 ha), increasing landscape fragmentation and forest edges. Efforts to quantify canopy texture and their link to carbon storage had limited to no success. Our analysis provides an insight in key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale and provide a historical context to land-cover and land-use change spatial patterns for past and ongoing field research in the area.
keywords: |
  Aerial survey, data recovery, CNN, Deep Learning, SfM, Congo Basin
acknowledgement: |
  This research was supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE). KH acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668). I'm grateful for the support of Thales D'Haulleville volunteering his time in scanning the images.
authorcontributions: |
  K.H. conceived and designed the study, analyzed the data, prepared figures and tables, authored, reviewed and approved the final draft of the manuscript. T.d.H. scanned all image data. E.K. and T.d.H. provided plot based AGC estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A., J.V.D.B., H.V and L.W. reviewed the final manuscript.
conflictsofinterest: |
  The authors declare no conflict of interest. The founding sponsors had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, an in the decision to publish the results.
output: rticles::mdpi_article
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  library(orthodrc)
  library(raster)
  library(sf)
  library(rgdal)
  library(ggplot2)
  library(ggthemes)
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(landscapemetrics)

  # set kable NA values to blank
  options(knitr.kable.NA = '')

@@

  row.names(change_stats) <- c(
   "Forest",
   "  -  of which edges",
   "Regrowth > 1958 loss",
   "Loss > 2000",
   "Loss > 1958 (permanent)")

  # drop square km values per reviewer request
  change_stats <- change_stats[,2:3]

# this data will not be available in the github repo due to missing geo-eye data

# read in cnn forest mask
m <- raster("../data/geo-eye/geo-eye_forest_mask_cnn.tif")

# hansen
hans <- raster("../data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")

# cloud mask
cloud_mask <- st_read("../data/maps/geo_eye_cloud_mask.shp")

# read in Hansen map (get all forest loss before 2013)
hans <- raster("../data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")
hans <- crop(hans, extent(m))
loss <- (hans <= 11) + (hans == 0)
loss <- (loss > 1)

# mask clouds
m <- mask(m, cloud_mask, inverse = TRUE)
loss <- mask(loss, cloud_mask, inverse = TRUE)

ct <- crosstab(m, loss)
ct <- round(ct/sum(ct) * 100,2)
rownames(ct) <- c("non-forest","forest")

# read in shapefile location digitized in QGIS
loc <- st_read("../data/cnn/sample_locations.shp")
disturbed <- loc %>% select(type) %>% filter(type == "dist")
forest <- loc %>% select(type) %>% filter(type == "forest")

# forest cover map
r <- raster("../data/orthomosaic/yangambi_forest_mask_qa_resampled.tif")
r <- (r > 0.5) * 1
r[is.na(r)] <- 3

# calculate the stats for the whole raster
forest_stats <- as.data.frame(na.omit(freq(mask(r, forest))))
forest_stats$count <- round(forest_stats$count/sum(forest_stats$count) * 100,2)

disturbed_stats <- as.data.frame(na.omit(freq(mask(r, disturbed))))
disturbed_stats$count <- round(disturbed_stats$count/sum(disturbed_stats$count) * 100,2)

valstats <- cbind(disturbed_stats$count, forest_stats$count)
rownames(valstats) <- c("non-forest","forest")

# calculate the stats for the validation
# area only
validation_forest <- loc %>%
  filter(cnn == "validate",
         type == "forest") %>%
  select(type)

validation_non_forest <- loc %>%
  filter(cnn == "validate",
         type == "dist") %>%
  select(type)

forest_stats <- as.data.frame(na.omit(freq(mask(r, validation_forest))))
forest_stats$count <- round(forest_stats$count/sum(forest_stats$count) * 100,2)

# add a row if forest cover is 100% accurate
if(forest_stats$count[1] == 100){
  if(forest_stats$value == 1){
   forest_stats <- rbind(c(0,0),forest_stats)
  }
}

disturbed_stats <- as.data.frame(na.omit(freq(mask(r, validation_non_forest))))
disturbed_stats$count <- round(disturbed_stats$count/sum(disturbed_stats$count) * 100,2)

valstats_subset <- cbind(disturbed_stats$count, forest_stats$count)

acc_ct <- sum(diag(ct))/sum(ct)
acc_valstats <- sum(diag(valstats))/sum(valstats)
acc_valstats_subset <- sum(diag(valstats_subset))/sum(valstats_subset)

acc <- round(c(NA,acc_ct,NA,acc_valstats,NA,acc_valstats_subset)*100,2)
valstats <- cbind(ct, valstats, valstats_subset)
valstats <- rbind(valstats, acc)
rownames(valstats) <- c("non-forest","forest","")

```

@@

Tropical ecosystem services are severely impacted by deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015]. Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function. Strong fragmentation effects decreases the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018]. Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].

As such, anthropogenic factors through LULCC heavily affect forest structure and functioning [@fauset2017; @brinck2017; @didham1999]. Accurate estimates of LULCC and forest canopy structure are therefore imperative to estimate carbon emissions and other ecosystem services [@houghton2012; @hansen2013]. Remote sensing products have been key input as they provide accurate spatial information to help estimate carbon emissions [@mitchard2018; @hansen2013]. High resolution aerial images provide scientists tools to monitor forest extent, structure and carbon emissions as canopy texture is linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010]. Yet, these estimates are often limited in time to recent decades [@defries2002; @houghton2012; @hansen2013; @achard2014].

Historical estimates **of Land-Use and Land-Cover (LULC)**, in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988]. Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post world war II German [@nita2018]. Survey data across the African continent is less common, inaccessible or both. Some studies do exist, as @buitenwerf2012, @hudak1998 and @okeke2006 used aerial survey images to map vegetation changes in South African savannas and the Adulam Nature Reserve, respectively, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930's.

Despite the existence of large historical archives of aerial survey imagery for the central Congo Basin (Figure 1, Appendix Figure 3) as of yet no studies have valorized these data. The lack of a consistent valorization effort is unfortunate as the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015]. Given the impact of LULCC on the structure and functioning of central African tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover warrants more attention [@willcock2016].

Here, we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi, in the central Congo Basin, effectively linking the start of the antrophocene [@lewis2015] with current assessments. Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin. Past research in the region allows for thorough assessment of land-use and land-cover change from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.

We leverage Structure-from-Motion to generate a large orthomosaic of historical imagery and develop a Convolutional Neural Network based forest cover mapping approach, based upon a semi-supervised generated dataset extensively leveraging data augmentation. Our methodology aims to provides a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity. We further contextualize the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin. Our fast scalable mapping approach for historical aerial survey data, using limited supervised input, would further support long-term land-use and land-cover change analysis across the central Congo Basin.

@@

Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville). Black-and-White infrared images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between **09:00 and 11:00** local time. Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negative of 180 by 180 mm. Flights were flown at an average absolute altitude of ~5200 m above sea level, covering roughly 18 530 km^2^ at an approximate scale of 1/40 000. The use of the integrated autograph system ensured timely acquisition of pictures with a precise overlap (~1/3) between images. This large overlap between images together with flight parameters would allow post-processing, using stereographs, to create accurate topographic maps. Original data from this campaign are stored in the [Royal Museum for Central Africa](https://www.africamuseum.be/en) in Tervuren, Belgium (Figure 1).

```{r echo=FALSE, out.width='75%', fig.align="center", fig.cap="A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river."}
knitr::include_graphics('./figures/print_examples.jpg')
```

@@

```{r echo=FALSE, out.width='75%', fig.align="center", message=FALSE, warning=FALSE, fig.cap="Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study. The bounding box of the orthomosaic data presented in this study is shown as a rectangle (23x36 km). The outline of a recent high-resolution Geo-eye panchromatic image is shown as a dashed dark grey rectangle (10x10km). The location of various permanent sampling plots are shown as x, +, and open squares and triangles for the mixed, mono-dominant and edge plots respectively. The grey polygon delineates the current day Yangambi Man and Biosphere reserve. The inset, top right, situates the greater Yangabmi region (white rectangle) with the DR Congo. The full flight plan and details are shown in Appendix Figure 1 and 2"}
knitr::include_graphics('./figures/flight_paths.png')
```

@@

All selected images, covering the Yangambi area, were contact prints as original negatives of the prints were not available. Images were scanned at a resolution exceeding their original resolution (or grain) at the maximal physical resolution of an Epson A3 flatbed scanner (i.e. 2400 dpi or 160 MP per image) and saved as lossless tiff images. Data were normalized using contrast limited histogram equalization [@zuiderveld1994] with a window size of 32 and a clip limit of 1.5. Fiduciary marks were used to rectify and downsample the images into square 7700x7700 pixel images (~1200 dpi, 81 MP). This resulted in a dataset with digital images at a resolution that remained above the visible grain of the photographs, whilst the reduced image size facilitated easier file handling and processing speed.

@@

We calculated the orthomosaic using a low resolution point cloud and digital elevation map (DEM). Additional ground control points were provided to assist in the referencing of image and constrain the optimization routine used in the SfM algorithm. Ground control points consisted of permanent structures which could be verified in both old and new aerial imagery (i.e. ESRI World Imagery) **and consisted of corner points of build structures (e.g. a building, bridge or swimming pool etc.).** Although clouds were removed during the SfM routine we did not mask all clouds in the final orthomosaic to maximize forest coverage. The final scene was cropped, to provide consistent wall to wall coverage of the reconstructed scene. The orthomosaic was exported as a geotiff for further georeferencing in QGIS [@qgisdevelopmentteam2018] using the georeferencer plugin (version 3.1.9) and additional ESRI World Imagery high resolution reference data. We used 3^th^ degree polynomial and 16 ground control points to correct the final image. Ground control points, raw image data and final processed image are provided in addition to measures of uncertainty such as mean and median error across all ground control points. All subsequent analysis are executed on the final geo-referenced orthomosaic or subsets of it.

@@

### Classifying Land-Use and Land-Cover

#### *Model training*

We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands which **showed visible canopy cover loss**, fields and buildings. We used the Unet Convolutional Neural Net (CNN, @ronneberger2015) architecture implemented in Keras [@chollet2015] with an efficientnetb3 pre-processing backbone [@yakubovskiy2019] running on TensorFlow [@martinabadi2015] to train a binary classifier (i.e. forest or non-forested). Training data were collected from the orthomosaic by randomly selecting 513 pixel square tiles from **locations within** homogeneous forested or non-forested polygons in the historical orthomosaic **(Figure 5). Separate polygons were selected for training, testing and validation purposes. Validation polygons were sampled 300 times, while both testing and validation polygons were sampled at 100 random locations. Tiles extracted from locations close to the polygon border at times contained mixed cover types. Tiles with mixed cover types were removed from the list of source tiles (Table 1).** Homogeneous **source** tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3). We generated 5000 synthetic landscapes for training, while 500 landscapes were generated for both the validation and the testing datasets for a total of 6000 synthetic landscapes. In order to limit stitch line misclassifications, along the seams of mosaicked images, we created synthetic landscapes with different forest tiles to mimick forest texture transitions. We applied this technique to 10% of the generated synthetic landscapes **(across training, validation and testing data)**.

```{r echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE}

training <- c(300, 294)
testing <- c(100, 92)
validation <- c(100, 84)

source_img <- data.frame(training, testing, validation)
rownames(source_img) <- c("forest","non-forest")

source_img %>%
   kable(
    caption="Number of source tiles used for the generation of synthetic landscapes.",
    booktabs = TRUE,
    escape = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))

```

```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="orthomosaic", fig.cap="An example synthetic landscape, combining homogeneous forest and non-forest images into a patchy landscape using random gaussian field based masks. The left panel shows a combined synthetic landscape, while the right panel shows the corresponding forest (black) and non-forest (white) labels."}
knitr::include_graphics('./figures/synthetic_landscape.png')
```

The CNN model was trained for 100 epochs with a batch size of 30 using Adam optimization [@kingma2017], maximizing the Intersect-over-Union (IoU) using Sørensen–Dice and categorical cross-entropy loss functions. Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring. The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote **(> 50% agreement)** across overlapping areas to limit segmentation edge effects. **In addition we provide, raw pixel level classification agreement data for quality control purposes (see Data availability below).** We refer to Figure 4 for a synoptic overview of the full deep learning learning workflow.

```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="orthomosaic", fig.cap="A diagram of the deep learning workflow followed in training a binary forest / non-forest cover convolutional neural net U-Net model to generate our forest cover map."}
knitr::include_graphics('./figures/cnn_diagram.pdf')
```

#### *Model validation*

**We report the CNN accuracy based upon the IoU of our out-of-sample validation dataset of synthetic landscapes. In addition, we report confusion matrices for all pixels across the homogeneous validation polygons, as well as the training and testing polygons (see Figure 5). Furthermore, we used the first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair (Geo-Eye, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 08:55 GMT or 09:55 local time) to classify and assess the robustness of the CNN algorithm on contemporary remote sensing imagery with similar spectral and spatial characteristics. We used the Global Forest Change version 1.6 (GFC, tile 10N-020E) [@hansen2013] map data, capturing forest loss up to 2011, to quantify accuracy on downsampled CNN Geo-Eye classification results. Once more, we report the confusion matrix of between the GFC and CNN derived forest cover maps, masking clouds and cloud shadows. To summarize confusion matrices we report accuracy as:**

$Accuracy = \frac{(TP + TN)}{(TP + TN + FP + FN)}$

**in which TP, TN, FP, FN are True Positive, True Negative, False Positive and False Negative, respectivelly.**

### Characterizing long term change

To map long term land-use and land-cover change in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) map data [@hansen2013]. Using the **GFC** data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier. In our analysis we only included **GFC** pixels which recorded no **forest** loss throughout the whole 2000 - 2018 period. As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30m GFC resolution. We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river. Where the datasets overlap we provide summary statistics of deforestation, reforestation and afforestation.

@@

Surveys of old plantations show a large variation in AGC, depending on age and the crop type. For example, the AGC values varied from 168.67 to 86.55 g Mg C ha^-1^, for *Hevea brasiliensis* (rubber tree) and *Elaeis guineensis* (oil palm) plots respectively (@bustillo2018, personal communications). These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see **Appendix Table 1**). We therefore use both the estimates of old-regrowth and mixed forest to estimate AGC for regrowth. We did not have sufficient data to account for individual changes in AGC across plantations.

@@

Hufkens et al. (2019). A curated dataset of aerial survey images over the central Congo Basin, 1958. Zenodo: doi.org/10.5281/zenodo.3547767. All data not included in the latter repository can be found bundled with the analysis code as listed below. Proprietary datasets (i.e. Geo-Eye data) are not shared, but purchase order numbers allow for acquisition of these datasets to ensure reproducibility.
